Rationality, Evolution \& the Environment

### 2015-10-03 paper/paper.tex
reshuffle first sections

\usepackage{todonotes}

@@

\title{Rationality, Evolution \& the Environment}
\author{Paolo Galeazzi \& Michael Franke}
\date{current version: September 15th 2015}

@@

\begin{definition}[Rational choice from maximization of expected utility] Let $S$ be a (finite)
  set of states of the world and $A$ a (finite) set of actions. Given a probability
  distribution $P$ over $S$ and a utility function
  $u \mycolon S \times A \rightarrow \mathbb{R} $, an action $a^* \in A$ is a rational choice
  only if it maximizes the \emph{(subjective) expected utility}
  $\EU(a^*) = \sum_{s \in S} P(s) \ u(s, a^*)$: $a^* \in \text{argmax}_{a_i \in A} \EU(a^*)
  $. \todo{Give finite formulation? Everything we do here is finite}
\end{definition}

@@

It could be objected that, without additional constraints, the notion of rationality is likely
to collapse, as it seems possible to deem rational everything that is observed, given the
freedom to adjust beliefs and preferences freely. Of course, there are many ways in which
ascription of beliefs and preferences is constrained further by normative considerations of
rationality. \todo{speak about Dutch books here?} To begin with, if the DM's uncertainty is
about objective chance, it can be argued that subjective beliefs should conform (at least
roughly) to objective chance. Similarly, in certain situations it may make sense to postulate
on normative grounds that subjective preferences should be oriented towards tracking objective
utility. For instance, in their classic work \myalert{(Alchian, Friedman)} argue that profit
maximization is a reasonable assumption for characterizing outcomes in a competitive market
because only firms behaving in a manner that is consistent with profit maximization will
survive in the long run.\todo{@Paolo: please add references}

@@

Taking inspiration from recent economic literature on the evolution of preference, we would
like to put this latter idea to the test. We

 Since the work by vNM [cite], this view has been justified from
the ``intuitive rationality'' of the axioms on the subjective preference relation leading to
the representability of the decision maker as an expected utility maximizer. \todo{include
  axioms, or at least those that are mentioned as relevant later} While in the axiomatization
by [vNM] the probability of the outcomes is objectively given a priori, Savage54 provides a
larger set of axioms that represent the decision maker (DM) with both a subjective utility over
outcomes \textit{and} a \textit{subjective} probability over states of the world. Savage's
axioms constitute the foundation of subjective expected utility (SEU) theory, that is normally
assumed to define the rational behavior of economic agents.

\begin{definition}[SEU-maximization] Let $S$ be a set of states of the world, $X$ a set of
  outcomes and $A$ a set of actions. Given a probability measure $P$ on $S$ and a subjective
  utility function $u:X \rightarrow \mathbb{R} $, an action $a^*: S \rightarrow X $ is
  SEU-maximizing if $a^* \in \text{argmax}_{a_i \in A} \int_S u(a^*(s))dP(s) $. \todo{do we
    actually need outcomes for anything in this paper?}
\end{definition}

The subjectivity of preferences plays a very important
role for the present work. From an agent's standpoint, the amount $x$ of, say, money achieved does not necessarily
coincide with the amount of subjective utility attained $u(x)$. 
Historically,  the first to notice and formally argue for this difference
between objective utility (goods, money, etc.) and subjective utility (preferences) was
D. Bernoulli with the famous St. Petersburg paradox in 1738. Moreover, Bernoulli introduced the idea that a good representation of
subjective utility can be carried out in terms of an increasing concave function of objective
utility. This can be considered the birth of modern decision theory.

Empirically, many violations of Savage's (and vNM's) axioms have been consistently observed in behavioral economics. Most of the attention has been focused on the Independence Axiom and the Sure-thing Principle. Allais' paradox and Ellsberg's paradox are examples of violations of these two axioms, and they are at the origin of Prospect theory [cite] and ambiguity in decision theory [cite].

The main idea behind Ellesberg's paradox and the literature on ambiguity is that DM treats differently objective and subjective uncertainty. If the probability measure on the possible states of the world is not given a priori and not ''known'', then observed preferences leave no room to represent  DM's beliefs via a probability measure on the possible states. The most common solution is then to represent DM's beliefs with a set of probability measures instead of a single one.

Prospect theory was introduced by Kahnemann and Tversky in
their famous paper from 1979. By means of experimental results they noticed that DM apparently acts as a risk averse agent with respect to gains and as a risk loving agent
with respect to losses. Consequently, they conjectured that the subjective utility is better
representable by an S-shaped utility function, concave in the positive domain of gains and
convex in the negative domain of losses.

More recently, other possible definitions of the subjective utility function have been proposed
and justified from a game theoretical point of view (e.g. Fehr and Schmidt, Charness and Rabin,
etc.). The central goal of these studies is to explain/describe the empirical behavior in social dilemmas
(e.g., Prisoner's dilemma, public goods game, etc.). In the same spirit of Prospect theory, by means of different subjective utility
functions it is possible to describe/match the behavior observed in game theoretical experiments and to
explain it as rational (in the sense of maximizing the subjective utility).
With this approach every behavior, no matter how far from the maximization of
objective utility, can be deemed rational if we can only find a subjective utility that
justifies that behavior as maximizing the subjective utility. The problem here is that it seems possible to consider rational everything
that is observed, given the adequate subjective utility function. The notion
of rationality almost collapses onto the observed behavior.

On the other side, other economists (Alchian, Friedman) seem to link the notion of rational choice
more tightly to the maximization of objective utility. In their classic works they argue that
profit maximization is a reasonable assumption for characterizing outcomes in a competitive
market because only firms behaving in a manner that is consistent with profit maximization will
survive in the long run. If the previous accounts of rationality looked more descriptive, this
definitely tastes more normative. In this sense, a normative notion of rationality would be tightly
related to (and almost coincide with) the maximization of objective utility, which is supposed
to drive the evolutionary process. It follows then that being normative in this sense points to an identification of rationality with maximization of objective utility.

It seems that we are stuck between two extrema: a too permissive descriptivism and a too restrictive normativism. On the one hand, if we want to have a
descriptive notion of rationality, then we can (almost) always define an appropriate
subjective utility whose maximization matches the observed behavior. On the other hand, if we
want to have a normative approach to rationality, then it seems we have to appeal to
evolutionary arguments entailing the understanding of rationality as maximization of objective
utility.

However, it has been argued that in a good theory of choice and decision making the normative
and the descriptive components should not come apart. For example, as Douglas Maclean [cite]
writes:

\todo[inline]{personally, not a big fan of extended quotes; talks: yes, papers: don't know; is
  this dear to your heart? who's that Maclean anyway? ;-)}

\begin{quote}
The history of the subject reflects strong reasons why the theory of choice should have this dual character. [...] \\
A normative theory of choice gains credibility from its predictive and explanatory power, for although empirical success itself may not constitute much of a normative argument, the theory of rational behavior is less likely to be found objectionable if rationality turns out to be widespread. [...] But if, on the other hand, the normative theory turns out to be incompatible with the best established behavioral theory, then the verification of the normative theory will depend entirely on the intrinsic plausibility of its axioms, and that appeal must be quite strong, strong enough to make us overcome our charitable reluctance to conclude that our actual behavior is systematically irrational.
\end{quote}

There is another violation of Savage's axioms that is particularly relevant for the present work: transitivity of preferences. The first of Savage's axioms states that the preference relation is a (total) preorder, i.e., reflexive and transitive. Moreover, the preorder axiom is often referred to as the rationality axiom [cite], as to suggest that it is somehow the core of rational preferences. Violations of transitivity have been observed and studied in experimental literature [citeTversky69,etc], but can also be supported from a normative and theoretical point of view. \\
In what follows, an important focus of our work will be the role of regret in decision making. Regret theory predicts the violation of both transitivity and Independence of Irrelevant Alternatives (IIA). [CONNECTING TO Fawcett\&al: this is IIA, not independence]

@@

Regret theory has been introduced in decision theory by Savage [] and Niherens [], and later developed by Bell [1982], Loomes and Sugden [1982], and Fishburn [1982] independently. Lately, Halpern and Pass [cite] show how the use of regret minimization can solve game theoretical puzzles (like the Centipede and the Traveller's dilemma) in a way that is closer to everyday intuition and empirical data. In this paper the notion of regret is defined as in [HalPass12]:

\begin{definition}[Pairwise regret] \label{defn:parwreg}
Given a simultaneous-move game $ G=\langle N, (A_i , \pi_i)_{i \in N} \rangle $, the regret of player $i$ at profile $(a_i,a_{-i})$ is defined by: $\text{reg}(a_i,a_{-i}):= \pi_i(a_i^\$,a_{-i})-\pi_i(a_i,a_{-i}) $, where $a_i^\$ \in A_i$ denotes $i$'s best reply to action $a_{-i} $.
\end{definition}

\begin{definition}[Regret] \label{defn:regret}
Given a simultaneous-move game $ G=\langle N, (A_i , \pi_i)_{i \in N} \rangle $, the regret of an action $a_i \in A_i$ is defined by: $\text{reg}(a_i):= \text{max}_{a_{-i}\in A_{-i}} \pi_i(a_i^\$,a_{-i})-\pi_i(a_i,a_{-i}) $.
\end{definition}

\begin{definition}[Regret minimization] \label{defn:regmin}
Given a simultaneous-move game $ G=\langle N, (A_i , \pi_i)_{i \in N} \rangle $, an action $a^{r}_i $ is regret minimizing if $a^{r}_i \in \text{min}_{a_i} \text{max}_{a_{-i}} \pi_i(a_i^\$,a_{-i})-\pi_i(a_i,a_{-i}) $.
\end{definition}

@@

\begin{figure}
\begin{center}%
\begin{tabular}{|c|c|c|}
\hline 
 & I & II\tabularnewline
\hline 
\hline 
I & 1;1 & 0;0\tabularnewline
\hline 
II & 0;0 & 2;2\tabularnewline
\hline 
\end{tabular}\end{center}

\protect\caption{A coordination game $G^C$}
\label{coordgame1}
\end{figure}

@@

The meta-game perspective that we introduce in this paper can shed more light on the duality of descriptivism and normativism, and opens
some space in between the two extrema. 
By keeping the evolutionary approach we want to retain a
normative perspective on rationality, but our results show that (under certain conditions)
evolution does not necessarily lead to identifying rationality with maximization of objective
utility, contrary to the intuition of Alchian [cite] and Friedman [cite].

The justification for this choice is both historical and behavioral. Among
the possible competitors of utility maximization, regret has always been in a prominent
position, both from a descriptive and normative viewpoint. Indeed, Bell82 writes:
\begin{quote}
  By explicitly incorporating regret, expected utility theory not only becomes a better
  descriptive predictor but also may become a more convincing guide for prescribing behavior to
  decision makers.
\end{quote}
In the same year, Graham Loomes and Robert Sugden \citet{LoomesSugden1982:Regret-Theory:-} show that a regret-based theory of choice can match
and explain all the main experimental results of prospect theory, and claim that
\begin{quote} [...] we shall challenge the idea that the conventional axioms [of decision
  theory] constitute the only acceptable basis for rational choice under uncertainty. We shall
  argue that it is no less rational to act in accordance with regret theory, and that
  conventional expected utility theory therefore represents an unnecessarily restrictive notion
  of rationality. \hfill \citep{LoomesSugden1982:Regret-Theory:-}
\end{quote}

@@

  \todo[inline]{I would like to suggest to move Figure 2 next to Figure 1; it is already useful
  for example 1 and could be referred to there already; I would suggest to use subfigures to
  put both next to each other}

@@

\begin{figure}
\begin{center}%
\begin{tabular}{|c|c|c|}
\hline 
 & I & II\tabularnewline
\hline 
\hline 
I & 0;0 & -2;-1\tabularnewline
\hline 
II & -1;-2 & 0;0\tabularnewline
\hline 
\end{tabular}\end{center}

\protect\caption{Regret representation $\tau^{reg}(G^C)$}
\label{coordgame1reg}
\end{figure}

@@

>>>>

\usepackage{todonotes}
\usepackage{subcaption}

@@

\title{Rationality, Evolution \& the Environment}
\author{Paolo Galeazzi \& Michael Franke}
\date{current version: October 3rd 2015}

@@

\begin{definition}[Rational choice from maximization of expected utility] Let $S$ be a (finite)
  set of states of the world and $A$ a (finite) set of actions. Given a probability
  distribution $P$ over $S$ and a utility function
  $u \mycolon S \times A \rightarrow \mathbb{R} $, an action $a^* \in A$ is a rational choice
  only if it maximizes the \emph{(subjective) expected utility}
  $\EU(a) = \sum_{s \in S} P(s) \ u(s, a)$: $a^* \in \text{argmax}_{a \in A} \EU(a)
  $.
\end{definition}

@@

It could be objected that, without additional constraints, the notion of rationality is likely
to collapse, as it seems possible to deem rational everything that is observed, given the
freedom to adjust beliefs and preferences freely. Of course, there are many ways in which
ascription of beliefs and preferences is constrained further by normative considerations of
rationality. \todo{speak about Dutch books here?} To begin with, if the DM's uncertainty is
about objective chance, it can be argued that subjective beliefs should conform (at least
roughly) to objective chance. When the DM's uncertainty cannot be anchored in an objective
notion of chance, then temporal consistency in his decision making should at least constrain
his subjective belief, else he fall prey to exploitation by Dutch books. Similarly, in certain
situations it may make sense to postulate on normative grounds that subjective preferences
should be oriented towards tracking objective utility, e.g., fitness. For instance, in their
classic work \myalert{(Alchian, Friedman)} argue that profit maximization is a reasonable
assumption for characterizing outcomes in a competitive market because only firms behaving in a
manner that is consistent with profit maximization will survive in the long run.\todo{@Paolo:
  please add references}

@@

Taking inspiration from recent economic literature on the evolution of preference, we would
like to put this latter idea to the test. We present a formal argument intended to support two
points: (i) subjective preferences that differ from objective fitness can be sustained and even
be adaptive; this is brought to the foreground if we acknowledge that (ii) it is important to
take into account the potential mutual dependence of subjective preferences and subjective
beliefs in adaptationist approaches to decision making. More specifically, we show that
subjective, non-veridical preference representations based on a notion of regret can be
adaptive and even outperform objective, veridical representations of the actual fitness, when
agents, at least on occasion, lack the ability to form reasonable probabilistic beliefs, in
which case they hold imprecise probabilistic beliefs (CITE). The following Section~XYZ will
introduce the basic notions needed for this argument, namely regret, imprecise probabilistic
beliefs and security strategies.

The method by which we want to advance our argument is to integrate ideas from the recent
literature on preference evolution (CITE) with basic tenets of adaptationism: we are interested
in studying the evolutionary competition of choice mechanism, conceived here as pairs of
(general ways of) subjective preference and subjective belief
representation. \textcolor{red}{EXAND}. This method is further detailed in
Section~\ref{sec:model}.

\section{Regret, imprecise probabilistic beliefs, choice mechanisms}

@@

\todo[inline]{this section starts off presupposing to much formal stuff for a philosophy
  audience; we should first say why we look at games, then give the definition of a game, then
  define regret, then example, ...}

Regret theory has been introduced in decision theory by Savage [] and Niherens [], and later
developed by Bell [1982], Loomes and Sugden [1982], and Fishburn [1982] independently. Lately,
\citet{HalpernPass2012:Iterated-Regret} showed how the use of regret minimization can solve
game theoretical puzzles (like the Centipede and the Traveller's dilemma) in a way that is
closer to everyday intuition and empirical data. In this paper the notion of regret is defined
as in \citet{HalpernPass2012:Iterated-Regret}:

\begin{definition}[Regret, regret minimization] \label{defn:regret} Given a simultaneous-move
  game $ G=\langle N, (A_i , \pi_i)_{i \in N} \rangle $, the regret of player $i$ at profile
  $(a_i,a_{-i})$ is $\text{reg}(a_i,a_{-i}):= \pi_i(a_i^\$,a_{-i})-\pi_i(a_i,a_{-i}) $, where
  $a_i^\$ \in A_i$ denotes $i$'s best reply to action $a_{-i} $. The regret of an action
  $a_i \in A_i$ is then
  $\text{reg}(a_i):= \text{max}_{a_{-i}\in A_{-i}} \pi_i(a_i^\$,a_{-i})-\pi_i(a_i,a_{-i}) $. An
  action $a^{r}_i $ is regret minimizing if
  $a^{r}_i \in \text{min}_{a_i} \text{max}_{a_{-i}} \pi_i(a_i^\$,a_{-i})-\pi_i(a_i,a_{-i}) $.
\end{definition}

@@

\begin{figure}

  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \begin{tabular}{ccc}
      \toprule
      & I & II \\
      \midrule
      I & 1;1 & 0;0 \\
      \hline 
      II & 0;0 & 2;2\\
      \bottomrule
    \end{tabular}
    \caption{Coordination game $G^C$}
    \label{coordgame1}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.4\textwidth}
    \centering
    \begin{tabular}{ccc}
      \toprule
      & I & II \\
      \midrule
      I & 0;0 & -2;-1 \\
      \hline 
      II & -1;-2 & 0;0\\
      \bottomrule
    \end{tabular}
    \caption{Regret representation $\tau^{reg}(G^C)$}
    \label{coordgame1reg}
  \end{subfigure}
  \caption{A coordination game (left) and its regret representation (right).}
    \label{coordgame1mainFig}
\end{figure}

@@

\todo[inline]{add results about regret: reg min is security choice on negative regret transform
(related to Figure 1b); regret choice equal to veridical in case of probabilistic beliefs and
maximization of expected gains}

\subsection{Imprecise probabilistic beliefs}
\label{sec:impr-prob-beli}

Consider the following problem. I have, on the desk in front of me, a bag containing either red or black marbles. There is no further information about the distribution of the two colors. What is then the probability that I will draw a red marble? As in Walley91-96, we are not talking about physical probability, but rather about epistemic probability, i.e., the subjective probability that an agent attributes to an event.
In the literature there are two traditional ways of answering this kind of questions. According to standard Bayesianism, one should always be able to determine a precise betting rate (that immediately translates into a precise probability distribution), which is fair, in the sense that the DM is indifferent between betting on or against red at the fair rate, and would bet on red at any rate that is more favourable than the fair rate and against red at any rate that is less favourable. In this case, a strict Bayesian, in the absence of any further information, would probably rely on the principle of insufficient reason and answer that the DM should have a uniform distribution that considers all the outcomes equipossible. \\
The second answer involves imprecise probabilities and is related to the literature about uncertainty, or ambiguity. Many authors argued in favor of this kind of approach [e.g., Walley91-96, Levi74, GardenforsANDSahlin82]. Since it is consistent with the information available that all the marbles are red, or that no marble is red, the resulting model should take it into account and specify a lower probability of red $\underline{P}(R)$ and an upper probability of red $\overline{P}(R)$ that define an interval of possible probabilistic beliefs [$\underline{P}(R), \overline{P}(R)$]. Given the information available in this case we would have $\underline{P}(R)=0$ and $\overline{P}(R)=1$. This approach appears particularly relevant and appropriate in a game theoretical context. Indeed, in a recent paper [Battigalliet al.15] Battigalli et al. write
\begin{quote}
Such uncertainty is inherent in situations of strategic interaction. This is quite obvious when such situations have been faced only a few times.
\end{quote}
In decision theory, the imprecise probability model is in line with most of the representation results about uncertainty [GilboaSchmeidler89,KMM06], and seems justified by empirical observations too (see below). Walley91-96 is a seminal work in the area, that also introduces learning procedures to get useful inferences and narrow down the probability set [$\underline{P}(R), \overline{P}(R)$] after observing data. Since we want to study the evolution of meta-types, the present work does not address the probem of drawing inferences from observations, and for the moment we do not endow the players with any learning procedure.\\
In line with traditional evolutionary game theory, we model a population of rather unsophisticated and myopic agents. %that possibly differ from each other with respect to two components: the player type and the epistemic type. The players 
The players encode both a psychological subjective representation of the interaction (player type) and a conceptualization of uncertainty (epistemic type), but they are too simple to have any learning or belief updating procedure. From this point of view, we can take the learning process away from natural selection and focus on player types and epistemic types as the target of evolutionary pressure.

@@

@@

@@

\begin{itemize}
\item The meta-game perspective that we introduce in this paper can shed more light on the duality of descriptivism and normativism, and opens
some space in between the two extrema. 
By keeping the evolutionary approach we want to retain a
normative perspective on rationality, but our results show that (under certain conditions)
evolution does not necessarily lead to identifying rationality with maximization of objective
utility, contrary to the intuition of Alchian [cite] and Friedman [cite].

The justification for this choice is both historical and behavioral. Among
the possible competitors of utility maximization, regret has always been in a prominent
position, both from a descriptive and normative viewpoint. Indeed, Bell82 writes:
\begin{quote}
  By explicitly incorporating regret, expected utility theory not only becomes a better
  descriptive predictor but also may become a more convincing guide for prescribing behavior to
  decision makers.
\end{quote}
In the same year, Graham Loomes and Robert Sugden \citet{LoomesSugden1982:Regret-Theory:-} show that a regret-based theory of choice can match
and explain all the main experimental results of prospect theory, and claim that
\begin{quote} [...] we shall challenge the idea that the conventional axioms [of decision
  theory] constitute the only acceptable basis for rational choice under uncertainty. We shall
  argue that it is no less rational to act in accordance with regret theory, and that
  conventional expected utility theory therefore represents an unnecessarily restrictive notion
  of rationality. \hfill \citep{LoomesSugden1982:Regret-Theory:-}
\end{quote}
\end{itemize}
